---
title: "Getting and Cleaning Data - Course Project"
author: "T. Porter "
date: "2024-01-17"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Getting and Cleaning Data - Course Project
## Codebook

# Introduction

This project generates a high-level summarized data set based on the summary datasets provided in the "Human Activity Recognition Using Smartphones" Version 1.0 dataset [1].  See the Data/UCI HAR Dataset/readme.txt file for details on how the summary datasets were produced.

The data was obtained from here:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

The raw data was downloaded and is now located within the "Data/UCI HAR Dataset" directory of this project.

The raw data was processed to give a tidy summary. The following shows how the data was processed to arrive at the final result.


# Project Output: Tidy Data Set (Data/datMeans.txt)
_The summarized tidy data set is located in the project Data folder and is named datMeans.txt_


Variable            | Domain | Description                     | Units    
--------------------|:------:|---------------------------------|----------
Activity | NA | The type of activity | Factor: WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING
DataType | NA | Indicates test or training data | Factor: Test, Train
tBodyAcc-mean()-XYZ | time| Mean of mean body acceleration measurements | g
tBodyAcc-std()-XYZ| time | Mean of body accel measurement standard deviation | g
tGravityAcc-mean()-XYZ | time | Mean of mean gravity accel measurements | g
tGravityAcc-std()-XYZ | time | Mean of gravity accel measurement std dev | g
tBodyAccJerk-mean()-XYZ | time | Mean of mean body jerk measurements | g/s 
tBodyAccJerk-std()-XYZ | time | Mean of body jerk measurement std dev | g/s
tBodyGyro-mean()-XYZ | time | Mean of mean body gyro measurements | rad/s
tBodyGyro-std()-XYZ | time | Mean of body gyro measurement std dev | rad/s
tBodyGyroJerk-mean()-XYZ | time | Mean of mean body gyro jerk measurements | rad/s^2
tBodyGyroJerk-std()-XYZ | time | Mean of body gyro jerk measurement std dev | rad/s^2
fBodyAcc-mean()-XYZ | freq | Mean of body accel mean freq domain measurements | Hz 
fBodyAcc-std()-XYZ | freq | Mean of body accel freq domain measurement std dev | Hz
fBodyAccJerk-mean()-XYZ | freq | Mean of body accel jerk mean freq domain measurements | Hz/s
fBodyAccJerk-std()-XYZ | freq | Mean of body accel jerk freq domain measurement std dev | Hz/s
fBodyGyro-mean()-XYZ | freq | Mean of body gyro mean freq domain measurements | Hz
fBodyGyro-std()-XYZ | freq | Mean of body gyro freq domain measurement std dev | Hz
|||

-------------------

# Data Processing Steps



**Step 0: Read in the raw data**

The data files used for this project are
```{r datapaths}
  pathDatTest    <- "../Data/UCI HAR Dataset/test/X_test.txt"     # test data
  pathNamesTest  <- "../Data/UCI HAR Dataset/test/y_test.txt"     # test activities
  pathDatTrain   <- "../Data/UCI HAR Dataset/train/X_train.txt"   # training data
  pathNamesTrain <- "../Data/UCI HAR Dataset/train/y_train.txt"   # training activities
  pathActivities <- "../Data/UCI HAR Dataset/activity_labels.txt" # activity labels
  pathFeatures   <- "../Data/UCI HAR Dataset/features.txt"        # varible names
```

The data files are read and stored for easy use
```{r readData}
  datTest    <- read.table(pathDatTest)
  datTrain   <- read.table(pathDatTrain)
  actTest    <- read.table(pathNamesTest)
  actTrain   <- read.table(pathNamesTrain)
  namesAct   <- read.table(pathActivities)
  namesFeat  <- read.table(pathFeatures)
```



**Step 1: Merge the training and test data**

Before merging the data, it is convenient at this time to first perform **Step 4: Label the dataset with descriptive names**.

```{r Step4}
  ### Step 1a & 4: Add descriptive variable names
  names(datTest)  <- namesFeat[,2]  # add names to test data
  names(datTrain) <- namesFeat[,2]  # add names to training data
```

**Step 1b: Create vectors containing "Test" or "Train" and add to datasets**

```{r Step1b}
  ## Create some vectors of "Test" and "Train"
  datTypeTest <- rep("Test", length(datTest[,1]))
  datTypeTrain <- rep("Train", length(datTrain[,1]))
  
  ## Insert them as first column in the datasets
  datTest  <- cbind(datTypeTest,  datTest)
  datTrain <- cbind(datTypeTrain, datTrain)
  
  ## Add a descriptive name
  names(datTest)[1]  <- "DataType"
  names(datTrain)[1] <- "DataType"
```

**Step 1c: Add the type of activity as the new first column**

```{r Step 1c}
  ##  Bind the activities as new first column (variable)
  datTest  <- cbind(actTest,  datTest)
  datTrain <- cbind(actTrain, datTrain)
  
  ## Add a column name
  names(datTest)[1]  <- "Activity"
  names(datTrain)[1] <- "Activity"
```

**Step 1d: Merge the test and training tables into a new table**

```{r Step1d}
  ## 1d. Merge the two tables into one
  datAll <- rbind(datTest, datTrain)
```

The new table now contains ALL the summary data. Activity types are still in code form; this will be addressed in Step 3.

At this point, our data looks like this:
```{r Step3Output, echo=FALSE}
  head(datAll)
```



**Step 2: Extract only the measurements on the mean and standard deviation for each measurement**

For the high-level summary output, We only want the variables associated with Means and Standard Deviations of the measurements.

_**NOTE:** Within the source data, there are two types of means:_
  _(1) Means of the time and frequency domain variables, and_ 
  _(2) Mean frequencies of the frequency domain variables_.

_For this project, the mean frequencies are not included. If the frequency domain means are needed, change the grep string to ".*mean.*"_

**Step 2a: Identify which columns to extract**

```{r Step2a}
  colAct  <- 1                                    ## 1st column (Activity)
  colType <- 2                                    ## 2nd column (Data Type)
  colMean <- grep(".*mean..-.*", names(datAll))  
  colStd  <- grep(".*std..-.*",  names(datAll))  
  
  # create sorted vector of column indices to keep
  colKeep <- names(datAll[sort(c(colAct, colType, colMean, colStd))])
```

**Step 2b: Extract the columns and store in a new table**

```{r Step2b1, echo=FALSE, results="HIDE"}
  ## needed for select()
  library(dplyr)
```

```{r Step 2b2}
  datAll <- datAll %>% select(all_of(colKeep))
```



**Step 3: Use descriptive activity names to describe the activity codes**

Verbose activity names are stored in the _namesAct_ table. 

```{r Step3}
  ## Replace the codes with the words in DatAll
  datAll$Activity <- namesAct[datAll$Activity, 2]
```



**Step 4: Add descriptive labels to dataset for variable names**

This was handled in Step 1, before the columns were extracted.
A detailed explanation of the variable names is given in the source data readme (Data/UCI HAR Dataset/readme.txt)



**Step 5: Create an independent tidy dataset with the average of each variable for each activity and each subject**

```{r Step5}
 datMeans <- datAll %>%                          ## datMeans is new tidy dataset
    group_by(Activity, DataType) %>%             ## set grouping
    summarize_at(vars(names(datAll)[-(1:2)]), mean)   ## Get means for all bu cols 1 & 2
```

The operation is complete. The datMeans table can be used as-is, or it can be written to disk.

```{r Step5Output, echo=FALSE}
  write.table(datMeans, "../Output/datMeans.txt")
  head(datMeans)
```

 
------------------------------------
**References:**
[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012